Supplemental Materials: Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva

### 2021-11-11 Reports/Paper1.Rmd
modifications to labeling in supplemental figures

```{r temp read n data}
basedir=here::here()
outdir=file.path(gsub('Code', 'Output', basedir), 'Figures')
datadir=gsub('Code', 'Data', basedir)
load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
load(paste0(datadir, '/CreatedData/allmethyl.Rdata'))
load(paste0(datadir, '/CreatedData/completeCaseSVA.Rdata'))
load(paste0(datadir, '/CreatedData/methylation_summarymodels.Rdata'))
```

```{r}
modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))%>%droplevels()
set_label(modeldata$childteen)='Visit'

n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
m_obs=modeldata%>%nrow()
n_twovisit=modeldata %>% group_by(id)%>%summarise(n=n())%>%filter(n==2)%>%nrow()

@@

```{r plotting variables}
theme_set(theme_bw())
set_label(modeldata$cm1bsex)='Child gender'
set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
set_label(modeldata$grim)='GRIM epigenetic clock (years)'
set_label(modeldata$globalmethylation)='Percent global DNA methylation'
set_label(modeldata$cg05575921)='AHRR gene: percent cg05575921 methylation'
set_label(modeldata$PostnatalMaternalSmokingAny)='Postnatal maternal smoking at ages 1 or 5'
set_label(modeldata$SmkAtVisitPastmonth)='Postnatal primary caregiver smoking in past month prior to visit'
```

@@

pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')))
corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)
dev.off()

@@

```{r supplemental correlation plot plot it , eval=as_supp, fig.width=12, fig.height=12}
corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8)

@@

outcome_labels=c('Global methylation', 'Pediatric clock', 'AHRR: cg05575921',  'Polymethylation score \n (newborn, mean centered)')

@@

forest_plot=all_models%>%
  mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
  unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes' & predictors=='base model')%>%
  filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(.~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.05)+
  theme_classic()+
  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))

ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=15)
main_fig=main_fig+1
```

@@

forest_plot_compareModels=all_models%>%
    mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
    unnest(tidied)%>%
    filter(term=='smkPreg_binaryYes')%>%
    filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
    geom_point(position = position_dodge(width = 1))+
    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
    facet_grid(predictors~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
    theme_classic()+
    theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
    scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))

@@

```{r surrogate variables heatmap}
#bring the heatmap function
covar=c('Sample_Plate', 'Slide', 'Array', 'cm1bsex', 'smkPreg_binary', 'childteen', 'ancestry')
modeldata$Sample_Plate=set_label(modeldata$Sample_Plate, 'Batch')
modeldata$cm1bsex=set_label(modeldata$cm1bsex, 'Child gender')
modeldata$PostnatalMaternalSmokingAny=set_label(modeldata$PostnatalMaternalSmokingAny, 'Postnatal maternal smoking (age 1 & 5)')
modeldata$SmkAtVisitPastmonth=set_label(modeldata$SmkAtVisitPastmonth, 'Postnatal maternal smoking (month prior to DNA methylation sample')

@@

plot_aprioriCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Fragile families: mixed effect model', 'Joubert et al. 2016 newborn cord blood', 'Joubert et al. 2016 peripheral blood older children', 'Wicklund et al. peripheral blood >16 years'))+scale_x_continuous('Estimate (95% CI)')

@@

fig4a= global_roc%>%
    dplyr::select(predictors, covariates, methylation, childteen, test, modeltype)%>%
    unnest(test) %>% 
    filter(str_detect(methylation, 'SSnewbornCT|cg0557|global') | modeltype=='Base model')%>%
    ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
    facet_wrap(childteen~., nrow=1, labeller=labeller(childteen=age_labels))+
    theme_classic()+
    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'), 
                       function(x) str_wrap(x, width = 30))+
    theme(legend.position = 'right')+guides(color=guide_legend(nrow=4, byrow=TRUE))+
    geom_abline(intercept=1, slope=-1, color='black')+
    xlab('Specificity')+ylab('Sensitivity')+
    theme(legend.position='bottom')+ coord_fixed()

>>>>

```{r temp read n data}
basedir=here::here()
outdir=file.path(gsub('Code', 'Output', basedir), 'Figures')
datadir=gsub('Code', 'Data', basedir)
load(paste0(datadir, '/CreatedData/allPhenoData.Rdata'))
load(paste0(datadir, '/CreatedData/allmethyl.Rdata'))
load(paste0(datadir, '/CreatedData/completeCaseSVA.Rdata'))
load(paste0(datadir, '/CreatedData/methylation_summarymodels.Rdata'))
modeldata=modeldata %>% mutate(childteen=factor(childteen, levels=c('Age 9', 'Age 15')))%>%droplevels()%>%copy_labels(modeldata)
```

```{r plotting variables}
theme_set(theme_bw())
set_label(modeldata$cm1bsex)='Child gender'
set_label(modeldata$cm1inpov)='Maternal income to poverty threshold (at baseline)'
set_label(modeldata$PostnatalMaternalSmokingDose)='Primary caregiver smoking past month by dose'
set_label(modeldata$SSnewbornCT_center)='Polymethylation score for smoke exposure'
set_label(modeldata$pediatric)='Pediatric epigenetic clock (years)'
set_label(modeldata$grim)='GRIM epigenetic clock (years)'
set_label(modeldata$globalmethylation)='Percent global DNA methylation'
set_label(modeldata$cg05575921)='AHRR gene: \n percent cg05575921 methylation'
set_label(modeldata$PostnatalMaternalSmokingAny)='Postnatal maternal smoking at ages 1 or 5'
set_label(modeldata$SmkAtVisitPastmonth)='Postnatal primary caregiver smoking in past month prior to visit'
set_label(modeldata$childteen)='Visit of DNA methylation measure'
```

```{r}
n_total=modeldata %>% dplyr::select(id)%>%unique()%>%nrow()
m_obs=modeldata%>%nrow()
n_twovisit=modeldata %>% group_by(id)%>%summarise(n=n())%>%filter(n==2)%>%nrow()

@@

@@

pdf(file.path(outdir, paste0('SupplementalFigure', sup_fig, '.pdf')), width=12)
corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8, number.cex= 7/ncol(res2))
dev.off()

@@

```{r supplemental correlation plot plot it , eval=as_supp, fig.width=12, fig.height=12}
corrplot.mixed(res2, upper="ellipse", lower='number', order='hclust', tl.pos ="lt", tl.cex=0.8, number.cex= 7/ncol(res2))

@@

outcome_labels=c('Percent global  \n DNA methylation  (percent)', 'Pediatric epigenetic \n clock (years)', 'Percent DNA methylation of \n cg05575921 in AHRR (percent)',  'Polymethylation score for prenatal maternal \n smoke exposure (standard deviations)')

@@

forest_plot=all_models%>%
  mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
  unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes' & predictors=='base model')%>%
  filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(.~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  geom_text(data=data_text, aes(x=x,label=n), position=position_dodge(width = 1), hjust=0.05)+
  theme_classic()+
  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
  scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Hispanic genetic ancestry', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))

ggexport(forest_plot, filename=file.path(outdir, paste0('Figure', main_fig, '.pdf')), width=19)
main_fig=main_fig+1
```

@@

forest_plot_compareModels=all_models%>%
    mutate(n=map_dbl(data, nrow), n=paste0('n=', n))%>%
    unnest(tidied)%>%
    filter(term=='smkPreg_binaryYes')%>%
    filter(outcome%in% my_outcomes)%>%
    mutate(outcome_f=factor(outcome, levels=rev(my_outcomes)))%>%
    ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=ancestry))+
    geom_point(position = position_dodge(width = 1))+
    geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
    facet_grid(predictors~outcome_f, scales='free_x', labeller = labeller(outcome_f=outcome_labels))+
    geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
    theme_classic()+
    theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
    scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')+
    scale_y_discrete('', limits=c('European ancestry', 'Hispanic ancestry', 'African ancestry', 'Unstratified'), labels=c('European genetic ancestry', 'Hispanic genetic ancestry', 'African genetic ancestry', 'Unstratified'))+scale_color_manual('', values=c('black', 'darkgrey'))

@@

```{r surrogate variables heatmap}
#bring the heatmap function
covar=c('Sample_Plate', 'Slide', 'Array', 'cm1bsex', 'smkPreg_binary', 'childteen', 'ancestry')
modeldata$Sample_Plate=set_label(modeldata$Sample_Plate, 'Sample Plate')
modeldata$cm1bsex=set_label(modeldata$cm1bsex, 'Child gender')
modeldata$PostnatalMaternalSmokingAny=set_label(modeldata$PostnatalMaternalSmokingAny, 'Postnatal maternal smoking (age 1 & 5)')
modeldata$SmkAtVisitPastmonth=set_label(modeldata$SmkAtVisitPastmonth, 'Postnatal maternal smoking (month prior to DNA methylation sample')

@@

plot_aprioriCpG<-ggplot(cpg_all, aes(x=estimate, xmin=conf.low, xmax=conf.high, color=source, y=CpG))+geom_point(position=position_dodge(width=1))+geom_errorbarh(height=0.1, position=position_dodge(width=1))+scale_x_continuous()+geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+facet_wrap(.~Gene, scales='free', ncol=2)+theme_classic()+theme(legend.position = 'bottom')+guides(color=guide_legend(nrow=4,byrow=TRUE))+geom_blank(data = blank_data, aes(x = estimate))+scale_color_discrete('', labels=c('Fragile families saliva: age 15', 'Fragile families saliva: age 9', 'Joubert et al. 2016 newborn cord blood', 'Joubert et al. 2016 peripheral blood older children', 'Wicklund et al. peripheral blood >16 years'))+scale_x_continuous('Estimate (95% CI)')

@@

fig4a= global_roc%>%
    dplyr::select(predictors, covariates, methylation, childteen, test, modeltype)%>%
    unnest(test) %>% 
    filter(str_detect(methylation, 'SSnewbornCT|cg0557|global') | modeltype=='Base model')%>%
  mutate(modeltype=gsub("Sustained smoking polymethylation score \\(newborns, cell-type controlled\\)", "Polymethylation score for prenatal maternal smoke exposure", modeltype))%>%
    ggplot(aes(x=spec, y=sens, color=modeltype))+geom_line()+
    facet_wrap(childteen~., nrow=1, labeller=labeller(childteen=age_labels))+
    theme_classic()+
    scale_color_manual('', values=c('red', 'darkred', 'darkblue', 'darkgreen', 'blue', 'green'), 
                       function(x) str_wrap(x, width = 30))+
    theme(legend.position = 'right')+guides(color=guide_legend(nrow=4, byrow=TRUE))+
    geom_abline(intercept=1, slope=-1, color='black')+
    xlab('Specificity')+ylab('Sensitivity')+
    theme(legend.position='bottom')+ coord_fixed()
